\malg: a novel interest management algorithm for distributed simulations of MMOGs

### 2008-05-01 trunk/textos/dsrt/engaoi.tex
eng
FINAL

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{\malg: a novell interest management algorithm for distributed simulations of MMGs}

@@

\begin{abstract}
%\boldmath
Traditionally, a central server is utilized to provide support to MMGs (massively multiplayer games), where the number of participants is in the order of tens of thousands. Much work has been done trying to create a fully peer-to-peer model to support this kind of application, in order to minimize the maintenance cost of its infrastructure, but critical questions remain. Examples of the problems relative to peer-to-peer MMG support systems are: vulnerability to cheating, overload of the upload links of the peers and difficulty to maintain consistency of the simulation among the participants. In this work, it is proposed the utilization of geographically distributed lower-cost nodes, working as a distributed server to the game. The distribution model and some related works are also presented. To address the communication cost imposed to the servers, we specify a novell refinement to the area of interest technique, significantly reducing the necessary bandwidth. Simulations have been made with ns-2, comparing different area of interest algorithms. The results show that our approach achieves the least bandwidth utilization, with a 33.10\% maximum traffic reduction and 33.58\% average traffic reduction, when compared to other area of interest algorithms.

@@

\section{Introduction}
% no \IEEEPARstart
In the last years, electronic games have become very popular, specially MMGs (massively multiplayer games), where the number of simultaneous players is in the order tens of thousands \cite{cecin2004fsa}. Worlf of Warcraft \cite{worldofwarcraft}, Lineage II \cite{lineage2} and Guild Wars \cite{guildwars} are successful examples of MMGs.

@@

There is another question, concerning the number of trasmissions that each peer needs to execute. In the client-server model, each client only needs to send his actions to the server, which processes them and broadcasts the resulting game state to every player. In the peer-to-peer model, each peer becomes responsible for processing its own player's actions and send the resulting state to everyone else to whom that state change might be relevant. The problem is that we cannot assume that every peer has enough bandwidth to do this. Besides, without a central server, the game becomes dependent on the peers' simulation, which can be corrupted in order to result in an invalid state which benefits a player, or even invalidate the whole game session.

@@

This paper is organized as follows: in section \ref{relatedwork}, some related works with proposed distribution models are presented; in section \ref{def}, some necessary definitions are made; in section \ref{model}, we propose a distribution model to be used as basis of our optimization; in section \ref{aoi}, we review some existing interest management algorithms; in section \ref{gaoi}, we present our approach in detail; in sections \ref{sim} and \ref{results}, the simulation and its results, respectively, are described and in section \ref{conclusion}, we make our final considerations.

@@

In the past few years, many works have been done trying to distribute the network support for massively multiplayer games. One approach is the peer-to-peer model, which has some issues concerning game state consistency among the the participating peers, vulnerability to cheating, and increased bandwidth usage by each peer. There are some proposals that try to minimize these problems. One of them is presented in \cite{schiele2007rpp}, where it is suggested the division of the virtual simulated game environment into regions, and for each region, a peer is chosen to be its coordinator. The function of this peer will be manage the players' interest. It will check who really needs each state update, so each player does not receive every update packet. This way, the peers' bandwidth usage is reduced. However, it will be still higher than with a server performing all the broadcasts. In the peer-to-peer model, each player usually must send updates to more than one other player. Besides, the chosen peer must be trustable to be the region coordinator.

@@

An example of this kind of model is described in \cite{assiotis2006dam}, where a distributed architecture for MMGs is proposed. It is also based on the division of the virtual environment into regions, each one with a server node assigned to it. The player whose avatar is situated in a region of the game world should connect to the server responsible for that region. This way, each server would group different players, based on their locality on the virtual space. To achieve consistency between the different server nodes performing the simulation, it is used the concept of locks. When a server node needs to alter the state of some entity of the game, it first needs to obtain exclusive access to that entity. To do that, it negociates with the other server nodes which might also need to make some change in the same object, and then obtains the lock. When it finishes changing the state, it releases the lock and notifies the other servers

@@

The first major restriction in the proposal of \cite{assiotis2006dam}, however, is the assumption that all server nodes are connected to one another through a high speed and low latency network, what cannot be assumed when using lower-cost geographically distributed nodes. Another problem is that the scalability question is addressed simply via the expansion of the virtual environment area, supposing that the players will spread over it. Finally, they suggest to solve the hotspot problem through successive recursive environment partitioning, until the number of players per server is below a certain threshold. However, there is a practical limit to the repartitioning of the virtual environment, and they do not suggest what to do when this limit is reached.

@@

This way, the details of this mechanism will not result in relevant changes to the interest management. When a player $P$, whose avatar is near the border of a region, performs actions whose results must be broadcasted to players with their avatars in other regions, the server $S$, responsible for $P$, simulates his actions, calculates the resulting state and simply sends it to the neighbor server, as if it was sending to its own clients. Similarly, when $S$ receives the resulting state of the action of a player who is connected to a neighbor server, it broadcasts the state to the clientes connected to it as if a player in its own region had executed that action.

@@

In the next sections, some versions of this technique, such as circular area based and avatar's field of view based interest management, will be presented. In the section \ref{gaoi}, it is introduced the approach of attenuating the frequency of updates, and a proposed algorithm will be described in detail.

@@

The simplest way to execute interest management is to define a circular area, whose center is determined by the coordinates of the avatar's location in the virtual environment. After that, the euclidian distance between the avatar and every other other entity in the game world is calculated. Let $A$ be an avatar, whose area of interest is a circle of radius $rad$. If an entity $E$ is at a distance less than $rad$ from $A$, then its state updates will be relevant to $A$. $A$ will not receive state updates from entities which are at a distance greater than $rad$. Figure \ref{fcirc} illustrates this type of area of interest.

@@

Before sending the state of an entity to a client, the last transmission time is checked. The next transmission is then scheduled to occur after a certain interval. If the relevance of that state is 1, the normal update interval will be used. If it is less than 1, the normal interval is divided by the relevance. For example, let the normal update interval of a game be \mbox{200 ms}. If the avatar $A_i$, which just sent a state update packet to $A_j$, is at a distance from $A_i$ such that its relevance is 0.5, the next transmission will only happen after an interval of 200 divided by 0.5. So, the player controlling $A_j$ will only receive an update of $A_i$ after 400 ms. Despite this interval is still less than a half second, it represents a reduction of the state update frequecy of $A_i$ in 50\%. As they are at a greater distance from one another, and the interval was increased only by 200 ms, this variation will probably be imperceptible by the player who controls $A_j$.

It is important to note that the reduction of the entities' update frequency may be combined with other interest management techniques. In \cite{boulanger2006cim}, various interest management algorithms are described, and they can be further improved if the idea of different transmission intervals based on the relevance of the updates is used. 
Generally the state of each entity is classified into one of only two extremes: it is relevant or it is not relevant, ignoring the fact that there is a wide range of intermediate values. The question is how to define the relevance value for each state update. In the next sections, it will be presented two examples of algorithms which define a method to determine this value, as well as the area of interest used by each one. In section \ref{csmooth}, it is specified the proposed \malg\ algorithm, which employs, besides other principles, the update frequency attenuation. The simulations and their results are shown in sections \ref{sim} and \ref{results}, respectively.

\subsection{Circle with attenuation}

A simple form of using various update intervals based on the relevance value would be utilizing the circular area of interest. To obtain the relevance of an entity relative to an avatar, the value may be set to 1 when they are in the same position and gradually reduce it as the distance between them increases, until it reaches 0 and stops decreasing no matter how further the entity goes. This is a way that, although simple, has shown a significant reduction in traffic between clients and servers. In Figure \ref{graded_circle}, it is illustrated how the area of interest would be, with entities' state update frequency attenuation for a given avatar.

@@

%           A                   A              A 
%\subsection{Ângulo de visão com área próxima e atenuação da freqüência de atualização}
\subsection{\malg\ algorithm}
%View Angle with close area and update frequency attenuation
%V    A          C     A         U      F         A
\label{csmooth}

The interest management algorithm proposed in this article, \malg\ (view angle with close area and update frequency attenuation), takes into account three main factors:

@@

As for entities that are outside the area nearby, but still within the avatar's field of view, their relevance must be calculated. It is proposed that the relevance of each entity decline gradually as the distance between it and the avatar in question increases. The further they are, the less frequent will be their state updates. This is possible because even if the interval update is doubled, it will most likely still be a fraction of a second, which is hardly noticeable for a player whose avatar is located at a great distance from the entity in question. Moreover, short delays between the arrival of state updates can be easily masked by extrapolation techniques, such as dead-reckoning \cite{smed2002rna}. The algorithm \ref{smalg} defines the operation of our interest management technique.

@@

>>>>

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{\malg: a novel interest management algorithm for distributed simulations of MMOGs}

@@

\begin{abstract}
%\boldmath

% :: OLD ABSTRACT ::
Traditionally, a central server is utilized to provide support to MMOGs (massively multiplayer online games), where the number of participants is in the order of tens of thousands. Much work has been done trying to create a fully peer-to-peer model to support this kind of application, in order to minimize the maintenance cost of its infrastructure, but critical questions remain. Examples of the problems relative to peer-to-peer MMOG support systems are: vulnerability to cheating, overload of the upload links of the peers and difficulty to maintain consistency of the simulation among the participants. In this work, it is proposed the utilization of geographically distributed lower-cost nodes, working as a distributed game server. The distribution model and some related works are also presented. To address the communication cost imposed to the servers, we specify the \malg\ algorithm, which is a novel refinement to the area of interest technique, significantly reducing the necessary bandwidth. Simulations have been made with ns-2 and their results demonstrate that out approach achieves the least bandwidth utilization, with a 33.10\% maximum traffic reduction and 33.58\% average traffic reduction, when compared to other area of interest algorithms.

% :: ABSTRACT++ BY FABIO ::
%Traditionally, a centralized server or cluster of server machines is employed to provide support for MMOGs (massively multiplayer online games), where the number of participants is in the order of tens of thousands. In order to minimize the maintenance cost of this infrastructure, much work has been done trying to create a fully decentralized or peer-to-peer model to support MMOGs, but some critical issues remain unsolved. Since there is no general availability of Internet-wide IP multicast routing, decentralization proposals usually require peers on home broadband links to upload real-time game traffic to several peers simultaneously, and widespread last-mile broadband technologies such as ADSL have low upstream capacities by design which are usually further capped by connectivity providers. This adds to the difficulty of maintaining the game simulation consistent among the peers, no matter how they are to be organized. In this work, we propose the use of geographically distributed lower-cost nodes, served by higher-end but still consumer-grade broadband links, which work together to act as a distributed MMOG serving entity. The distribution model and some related works are also presented. To address the communication cost imposed on those `super-peer' server nodes, we also propose a novel refinement to the area of interest technique which significantly reduces the bandwidth consumed by game update traffic. Simulations have been made with ns-2, comparing different area of interest algorithms. The results show that our approach achieves the least bandwidth utilization, with a 33.10\% maximum traffic reduction and 33.58\% average traffic reduction, when compared to other area of interest algorithms.

@@

\section{Introduction}
% no \IEEEPARstart
In the last years, electronic games have become very popular, specially MMOGs (massively multiplayer online games), where the number of simultaneous players is in the order tens of thousands \cite{cecin2004fsa}. World of Warcraft \cite{worldofwarcraft}, Lineage II \cite{lineage2} and Guild Wars \cite{guildwars} are successful examples of MMOGs.

@@

There is another question, concerning the number of transmissions that each peer needs to execute. In the client-server model, each client only needs to send his actions to the server, which processes them and broadcasts the resulting game state to every player. In the peer-to-peer model, each peer becomes responsible for processing its own player's actions and send the resulting state to everyone else to whom that state change might be relevant. The problem is that we cannot assume that every peer has enough bandwidth to do this. Besides, without a central server, the game becomes dependent on the peers' simulation, which can be corrupted in order to result in an invalid state which benefits a player, or even invalidate the whole game session.

@@

This paper is organized as follows: in section \ref{relatedwork}, some related works with proposed distribution models are presented; in section \ref{def}, some necessary definitions are made; in section \ref{model}, we propose a distribution model to be used as basis of our optimization; in section \ref{aoi}, we review some existing interest management algorithms; in section \ref{gaoi}, we explain the principle of using variable update intervals; in section \ref{csmooth}, we present in detail the \malg\ algorithm; in sections \ref{sim} and \ref{results}, the simulation and its results, respectively, are described and in section \ref{conclusion}, we make our final considerations.

@@

In the past few years, many works have been done trying to distribute the network support for massively multiplayer online games. One approach is the peer-to-peer model, which has some issues concerning game state consistency among the participating peers, vulnerability to cheating, and increased bandwidth usage by each peer. There are some proposals that try to minimize these problems. One of them is presented in \cite{schiele2007rpp}, where it is suggested the division of the virtual simulated game environment into regions, and for each region, a peer is chosen to be its coordinator. The function of this peer will be manage the players' interest. It will check who really needs each state update, so each player does not receive every update packet. This way, the peers' bandwidth usage is reduced. However, it will be still higher than with a server performing all the broadcasts. In the peer-to-peer model, each player usually must send updates to more than one other player. Besides, the chosen peer must be trustable to be the region coordinator.

@@

An example of this kind of model is described in \cite{assiotis2006dam}, where a distributed architecture for MMOGs is proposed. It is also based on the division of the virtual environment into regions, each one with a server node assigned to it. The player whose avatar is situated in a region of the game world should connect to the server responsible for that region. This way, each server would group different players, based on their locality on the virtual space. To achieve consistency between the different server nodes performing the simulation, it is used the concept of locks. When a server node needs to alter the state of some entity of the game, it first needs to obtain exclusive access to that entity. To do that, it negotiates with the other server nodes which might also need to make some change in the same object, and then obtains the lock. When it finishes changing the state, it releases the lock and notifies the other servers

@@

The first major restriction in the proposal of \cite{assiotis2006dam}, however, is the assumption that all server nodes are connected to one another through a high speed and low latency network, what cannot be assumed when using lower-cost geographically distributed nodes. Another problem is that the scalability question is addressed simply via the expansion of the virtual environment area, supposing that the players will spread over it. Finally, they suggest to solve the hot spot problem through successive recursive environment partitioning, until the number of players per server is below a certain threshold. However, there is a practical limit to the repartitioning of the virtual environment, and they do not suggest what to do when this limit is reached.

@@

This way, the details of this mechanism will not result in relevant changes to the interest management. When a player $P$, whose avatar is near the border of a region, performs actions whose results must be broadcasted to players with their avatars in other regions, the server $S$, responsible for $P$, simulates his actions, calculates the resulting state and simply sends it to the neighbor server, as if it was sending to its own clients. Similarly, when $S$ receives the resulting state of the action of a player who is connected to a neighbor server, it broadcasts the state to the clients connected to it as if a player in its own region had executed that action.

@@

In the next sections, some versions of this technique, such as circular area based and avatar's field of view based interest management, will be presented. In section \ref{gaoi}, it is introduced the approach of attenuating the frequency of updates, and in section \ref{csmooth} our proposed algorithm will be described in detail.

@@

The simplest way to execute interest management is to define a circular area, whose center is determined by the coordinates of the avatar's location in the virtual environment. After that, the euclidian distance between the avatar and every other entity in the game world is calculated. Let $A$ be an avatar, whose area of interest is a circle of radius $rad$. If an entity $E$ is at a distance less than $rad$ from $A$, then its state updates will be relevant to $A$. $A$ will not receive state updates from entities which are at a distance greater than $rad$. Figure \ref{fcirc} illustrates this type of area of interest.

@@

Before sending the state of an entity to a client, the last transmission time is checked. The next transmission is then scheduled to occur after a certain interval. If the relevance of that state is 1, the normal update interval will be used. If it is less than 1, the normal interval is divided by the relevance. For example, let the normal update interval of a game be \mbox{200 ms}. If the avatar $A_i$, which just sent a state update packet to $A_j$, is at a distance from $A_i$ such that its relevance is 0.5, the next transmission will only happen after an interval of 200 divided by 0.5. So, the player controlling $A_j$ will only receive an update of $A_i$ after 400 ms. Despite this interval is still less than a half second, it represents a reduction of the state update frequency of $A_i$ in 50\%. As they are at a greater distance from one another, and the interval was increased only by 200 ms, this variation will probably be imperceptible by the player who controls $A_j$.

It is important to note that the reduction of the entities' update frequency may be combined with other interest management techniques. In \cite{boulanger2006cim}, various interest management algorithms are described, and they can be further improved if the idea of different transmission intervals based on the relevance of the updates is used. 
Generally the state of each entity is classified into one of only two extremes: it is relevant or it is not relevant, ignoring the fact that there is a wide range of intermediate values. The question is how to define the relevance value for each state update.

% \subsection{Circle with attenuation}

A simple form of using various update intervals based on the relevance value would be utilizing the circular area of interest. To obtain the relevance of an entity relative to an avatar, the value may be set to 1 when they are in the same position and gradually reduce it as the distance between them increases, until it reaches 0 and stops decreasing no matter how farther the entity goes. This is a way that, although simple, has shown a significant reduction in traffic between clients and servers. In Figure \ref{graded_circle}, it is illustrated how this area of interest would be, with entities' state update frequency attenuation for a given avatar.

@@

In the next section, it will be presented the proposed \malg\ algorithm, which employs, besides other principles, the update frequency attenuation. The simulations and their results are shown in sections \ref{sim} and \ref{results}, respectively.

%           A                   A              A 
%\subsection{Ângulo de visão com área próxima e atenuação da freqüência de atualização}
\section{The \malg\ algorithm}
%View Angle with close area and update frequency attenuation
%V    A          C     A         U      F         A
\label{csmooth}

In the previous sections, we have hinted that filtering based on an avatar field of view of 180 degrees (for instance) instead of using a full circle around it could result in significant reduction in update traffic from the server to the avatar player's machine. However, that would bring an important drawback, which is the large but temporary inconsistencies that would ensue whenever a player would turn around his avatars quickly to see what remotely-controlled objects are behind it. We have also reminded that reducing update frequencies for objects that are farther away from an avatar is a good optimization, especially for large virtual worlds where it may be on an open field with potentially hundreds of other avatars in its sighting distance. In this case, the player will not benefit at all from receiving updates at maximum frequency from objects that are sufficiently distant.
%from using the maximum update frequency that is granted to objects that are at maximum closeness to the player being updated.

In our envisioned distributed MMOG scenario, the field of view optimization cannot be missed. We need to use the avatar's field of view information to further improve on filtering gains. Furthermore, as it has the mentioned drawback, we combined it with a smaller circle around the avatar. %To achieve this, we propose the \malg\ (view angle with close area and update frequency attenuation) algorithm, which combines a circle area with an area based on the field of view of the avatar, as seen in Figure \ref{fcsmooth}.
This addresses the field of view blind spot because, even if the game allows players to turn their avatars 180 degrees quasi-instantly, the only objects which will be affected negatively due to abrupt turning will be the more distant ones. %This is the case of most 3D action games such as Quake \cite{quake}, in which distant objects are already less relevant for e.g. aiming and shooting purposes.
 Some time will be taken for the client to inform the server about the new facing angle of its avatar and to receive the state updates of the entities which were far behind it. Nevertheless, this delay will affect only distant objects, what is acceptable for most games, even action or shooter ones, which are the most demanding in terms of consistency due to presence of e.g. long-range hit-scan weapons.

%The time that it takes for the client to inform the server about the new facing angle of its avatar and to receive the next batch of object updates will be the size of the temporal inconsistency for distant objects behind that avatar, which is acceptable for most games, even action or shooter games which are the most demanding in terms of consistency due to the presence of e.g. long-range hit-scan weapons.

Therefore, the interest management algorithm proposed in this article, \malg\ (view angle with close area and update frequency attenuation), takes into account three main factors:

@@

As for entities that are outside the area nearby, but still within the avatar's field of view, their relevance must be calculated. It is proposed that the relevance of each entity decline gradually as the distance between it and the avatar in question increases. The farther they are, the less frequent will be their state updates. This is possible because even if the interval update is doubled, it will most likely still be a fraction of a second, which is hardly noticeable for a player whose avatar is located at a great distance from the entity in question. Moreover, short delays between the arrival of state updates can be easily masked by extrapolation techniques, such as dead-reckoning \cite{smed2002rna}. The algorithm \ref{smalg} defines the operation of our interest management technique.

@@

Of course, using field of view based filtering, there will be always some kind of impact, even if just an aesthetic one. However, for our proposed MMOG distribution scheme, with the kind of connectivity we are aiming at, this is a no-issue, since bandwidth savings at the servers not only help the game to scale but also enables the architecture to actually work by lowering the barrier of entry of nodes that wish to volunteer as servers in the proposed MMOG network.
